---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cwcens

<!-- badges: start -->
<!-- badges: end -->

This package implements non-parametric estimation via a kernel estimator 
for the probability in state and restricted mean time in state in an
illness-death model under component-wise censoring. Component-wise censoring
arises when illness can only be measured at a finite set of times, while death
is right censored and thus observed continuously up to the right censoring time.
Component-wise censored composite endpoints arise often in biostatistical
practice. For example, in many oncology studies, progression-free survival is
component-wise censored. Many articles in the literature describe the bias and 
"bumpy" Kaplan-Meier curves that can result from applying standard survival 
methods to component-wise censored data. 

## Installation

You can install the development version of cwcens from [GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
devtools::install_github("anneae/cwcens")
```
## Irreversible Illness-Death Model Example 

First, we will simulate data from an irreversible illness-death model. 

```{r, echo = F}
states <- c("State 1\n Alive without \n illness", 
            "State 2\n Alive with \n illness", 
            "State 3\n Dead")
connect <- matrix(0, 3, 3, dimnames=list(states, states))
connect[1, 2:3] <- 1; connect[2,3]<-1
survival::statefig(c(1, 2), connect)
```

Our dataset will contain 200 patients. Visits are scheduled every six months up 
to four years, and actual visit times are scattered around the scheduled visit
times according to a normal distribution with a standard deviation of 10 days.

```{r}
library(cwcens)

irrevdat <- simdat(200, scale12=1/.0008, scale13=1/.0002, scale23=1/.0016,
       scale21=1/.0008, vital.lfu=c(30.4*36, 30.4*48),
       visit.schedule = 30.4*c(12, 24, 36, 48), scatter.sd=10, 
       seed = 123)

head(irrevdat)
```

The dataset `irrevdat` records the time that each visit occurred and the 
individual's current state at each visit. For example, the individual in the
first row of `irrevdat` had visits at `r round(irrevdat$t1[1],2)`, `r round(irrevdat$t2[1],2)`
and `r round(irrevdat$t3[1],2)` days, and their state at those visits was 
`r round(irrevdat$x1[1],2)`, `r round(irrevdat$x2[1],2)` and `r round(irrevdat$x3[1],2)`,
respectively, meaning they were still in state 1 (alive and illness-free) at 
the third visit. The individual was right censored at time `r round(irrevdat$dtime[1],2)`.

We can estimate the probability of being in each state at 4 years with 
the following code:
```{r}
PIS<-kernel.est(irrevdat, bandwidth = 30.4*12, tau2 = 30.4*48, prob.times = 30.4*48,
                boundary = 'interpolation')
PIS
```

We estimate that an individual will be alive and illness-free at 4 years with probability `r round(PIS$prob.info[2], 3)`, alive with illness with probability `r round(PIS$prob.info[3], 3)`, and dead with 
probability `r round(PIS$prob.info[4], 3)`.

We can estimate the restricted mean time in each state in the first 4 years with 
the following code:
```{r}
RMTISdays<-kernel.est(irrevdat, bandwidth = 30.4*12, tau2 = 30.4*48, mu.times = 30.4*48,
                boundary = 'interpolation')
RMTISdays
RMTISyears<-kernel.est(irrevdat, bandwidth = 30.4*12, tau2 = 30.4*48, mu.times = 30.4*48,
                boundary = 'interpolation', scale=12*30.4)
RMTISyears
```

We estimate that an individual will spend `r round(RMTISyears$mu.info[2],2)` years (`r round(RMTISdays$mu.info[2],0)` days) in state 1, `r round(RMTISyears$mu.info[3],2)` years (`r round(RMTISdays$mu.info[3],0)` days) in state 2, and `r round(RMTISyears$mu.info[4],2)` years (`r round(RMTISdays$mu.info[4],0)` days) in state 3, on average, out of the first 4 years. 

Below, we estimate the probability of being in each state over time (specifically, each day up to 1459 days or 4 years), and plot the results.

```{r}
pmat<-kernel.est(irrevdat, bandwidth = 30.4*12, tau2 = 30.4*48, prob.times = 1:1459, boundary = 'interpolation')
head(pmat$prob.info)
plot((1:1459)/365.25, pmat$prob.info$p1, type = 'l', col = 'dark green', 
     xlab = 'Years', ylab = 'Probability in State', ylim = c(0,1))
lines((1:1459)/365.25, pmat$prob.info$p2, col = 'goldenrod')
lines((1:1459)/365.25, pmat$prob.info$p3, col = 'red')
```
The green, yellow and red lines represent the probability of being alive and
illness-free, alive with illness, and dead, over time. 

Note that the green curve represents the probability of not having experienced 
a composite endpoint consisting of death and illness. A common approach for estimating the illness-free survival probability in this setting is to apply
the Kaplan-Meier estimator to the observed data. That is, treat the first time 
illness is *observed* at a visit as the date it actually occurred, and if
illness was not observed at a visit, assume illness did not occur.  Calculate
the (possibly right censored) time to the earlier of illness or death for each
person and apply the Kaplan-Meier estimator. The following code implements
this approach and plots the resulting curve (green dotted line) along with the curve
estimated with the kernel estimator (green solid line) and the true probability of being 
alive and illness-free based on the simulation parameters (black solid line). 

```{r}
library(survival)
etime <- ifelse(irrevdat$state2obs<Inf, irrevdat$state2obs, 
                ifelse(irrevdat$dstatus==1, irrevdat$dtime, irrevdat$laststate1))
event <- ifelse(irrevdat$state2obs<Inf, 1, irrevdat$dstatus)

kmest <- survfit(Surv(etime, event)~1)
kmest

plot((1:1459)/365.25, pmat$prob.info$p1, type = 'l', col = 'dark green', 
     xlab = 'Years', ylab = 'Probability in State', ylim = c(0,1))
lines(stepfun(kmest$time/365.25, c(1,kmest$surv)), col = 'dark green', 
      do.points= F, lty = 3)
lines((1:1459)/365.25, pexp((1:1459), rate = 0.0002+0.0008, lower.tail = F))
```

The standard approach leads to a curve with noticeable drops around the time of 
scheduled visits (i.e. at 1, 2, 3 and 4 years) since illness is first *observed* 
around those times in many patients. In addition, when using the standard approach, 
we right censored people without observed death or illness at the time of last 
visit. This means that the number at risk is low after ~3 years, and it forces 
the survival curve to drop to zero over the observed followup period. 
Meanwhile, the kernel approach yields more plausible estimates. The kernel approach 
takes into account the fact that visits offer a "snapshot" of current illness 
status and the associated uncertainty about the exact time that illness developed.

## Reversible Illness-Death Model Example

Next, we will simulate and analyze data from a reversible illness-death model. 
In a reversible model, a patient's illness can resolve, so transitions 
back to the illness-free state (state 1) from the illness state (state 2) are allowed.

```{r, echo = F}
states <- c("State 1\n Alive without \n illness", 
            "State 2\n Alive with \n illness", 
            "State 3\n Dead")
connect <- matrix(0, 3, 3, dimnames=list(states, states))
connect[1, 2:3] <- 1; connect[2,3]<-1
connect[2, 1] <- 1
survival::statefig(c(1, 2), connect)
```

When the `scale21` parameter in the `simdat` function is greater than zero, 
data is simulated under a reversible model.

```{r}
revdat <- simdat(150, scale12=1/.0008, scale13=1/.0001, scale23=1/.0008,
       scale21=1/.0016, vital.lfu=c(30.4*36, 30.4*48),
       visit.schedule = 30.4*c(12, 24, 36, 48), scatter.sd=10, 
       seed = 111)

head(revdat)
```

The individual in the
first row of `revdat` had visits at `r round(revdat$t1[1],2)`, `r round(revdat$t2[1],2)`
and `r round(revdat$t3[1],2)` days, and their state at those visits was 
`r round(revdat$x1[1],2)`, `r round(revdat$x2[1],2)` and `r round(revdat$x3[1],2)`,
respectively, meaning that the patient developed the illness
between the first and second visit, and the illness cleared up
between the second and third visit. 
The individual was then right censored at time `r round(revdat$dtime[1],2)`.

We next estimate and plot the probability of being in each state over time:
```{r}
pmat<-kernel.est(revdat, bandwidth = 30.4*12, tau2 = 30.4*48, prob.times = 1:1459, boundary = 'interpolation')
head(pmat$prob.info)
plot((1:1459)/365.25, pmat$prob.info$p1, type = 'l', col = 'dark green', 
     xlab = 'Years', ylab = 'Probability in State', ylim = c(0,1))
lines((1:1459)/365.25, pmat$prob.info$p2, col = 'goldenrod')
lines((1:1459)/365.25, pmat$prob.info$p3, col = 'red')
```

Finally, we will use the Aalen-Johansen estimator for multistate models, implemented
in the `survival` package, and compare the results to the results from the kernel 
estimator. Like the Kaplan-Meier estimator, the Aalen-Johansen estimator is not intended for component-wise censored data. With component-wise censoring, we don't observe the
actual times of transitions between states. If a person is in one state at a given 
visit, and in a different state at the next visit, we only know a transition occurred
sometime between those two visits. To apply the Aalen-Johansen estimator, we will 
assume that the transition actually occurred at the time of the second visit, 
when the new state was first observed. The following code makes a long dataset with 
counting process structure, which will be the input when we use `survfit` to 
get the Aalen-Johnasen estimates. 

```{r}
revdat$id<-1:nrow(revdat)
revdat_long <- data.frame(id = revdat$id, tstart=0, 
                          tstop = ifelse(!is.na(revdat[,'t1']), 
                                         revdat[,'t1'], revdat$dtime),
                          event = ifelse(!is.na(revdat[,'t1']), 
                                         revdat[,'x1'], revdat$dstatus*3))
for(j in 2:revdat$nvisits[1]){
  revdat_long <- rbind(revdat_long, 
          data.frame(id = revdat$id, tstart=revdat[,paste('t',j-1,sep ='')], 
          tstop = ifelse(!is.na(revdat[,paste('t',j, sep='')]),  
                        revdat[,paste('t',j, sep='')],
                        ifelse(revdat$dstatus==1, 
                               revdat$dtime, revdat[,paste('t',j-1,sep ='')])),
          event = ifelse(!is.na(revdat[,paste('t',j, sep='')]),  
                        revdat[,paste('x',j, sep='')],revdat$dstatus*3)))
}

revdat_long<-rbind(revdat_long, data.frame(id = revdat$id, 
                tstart= revdat[,paste('t', revdat$nvisits[1], sep ='')],
                tstop = ifelse(revdat$dstatus==1, revdat$dtime,
                              revdat[,paste('t',revdat$nvisits[1], sep ='')]) , 
                event = revdat$dstatus*3))

revdat_long <- revdat_long[!is.na(revdat_long$tstart) & revdat_long$tstop > revdat_long$tstart, ]
revdat_long <- revdat_long[order(revdat_long$id, revdat_long$tstart),]
revdat_long$event<- factor(revdat_long$event, 0:3, labels=c("censor", "(s0)","st2", "st3"))
    
head(revdat_long)
```

Below, we calculate the Aalen-Johansen estimates of the probability in state and 
restricted mean time in state, and plot the probability in state estimates along 
with the estimates from the kernel estimator. The green, yellow and red lines correspond to states 1, 2 and 3, and the dotted lines represent the Aalen-Johansen estimator 
while the solid lines represent the kernel estimator.  
```{r}
ajest <- survfit(Surv(tstart, tstop, as.factor(event))~1, id = id, dat = revdat_long, influence=T)
ajest

plot((1:1459)/365.25, pmat$prob.info$p1, type = 'l', col = 'dark green', 
     xlab = 'Years', ylab = 'Probability in State', ylim = c(0,1))
lines((1:1459)/365.25, pmat$prob.info$p2, col = 'goldenrod')
lines((1:1459)/365.25, pmat$prob.info$p3, col = 'red')
lines(stepfun(ajest$time/365.25, c(1,ajest$pstate[,1])), 
      col = 'dark green', do.points = F, lty = 3)
lines(stepfun(ajest$time/365.25, c(0,ajest$pstate[,2])), 
              col = 'goldenrod',  do.points = F, lty = 3)
lines(stepfun(ajest$time/365.25, c(0,ajest$pstate[,3])), 
              col = 'red', do.points = F, lty = 3)

```


